Liveness-Based Garbage Collection for Lazy Languages

### 2015-07-11 paper/paper.tex
lemma ref.
lemma

\begin{figure}[t]
\renewcommand{\arraystretch}{1.2}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\, e_\mainpgm
    \hspace{5em} \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

\subsection{Semantics}
We now give  a small-step semantics for our  language.
We first specify the domains used by the semantics:
\[
\renewcommand{\arraystretch}{1.1}
\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
\end{array}
\]

@@

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
%==============================================================
\begin{figure*}[t]
\renewcommand{\arraystretch}{1.5}
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
$\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha
\mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne
\emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
\mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
\begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{\epath} \mapsto
\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{\epath}$ is a new e-path
  terminating with $\psi$}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) =
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
\ldots \mathit{df_k}$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
  on the expression $e$, and $e_f$ is the body of the function. As the
  section on liveness  analysis shows, The entire  function body $e_f$
  and $\sigma$  are needed  to calculate  the liveness  environment at
  different program  points. The  demand sigma  is ``dynamic''  in the
  sense  that it  arises from  the actual  function calls  made during
  execution.   The  0-CFA  demand  is an  over-approximation  of  this
  demand.  The information  in a suspended evaluation  context is also
  similarly augmented with  the demand on it and the  function body of
  which the suspended evaluation is a  part.  Thus a stacked entry now
  takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
  small step semantics which carries the extra information is shown in
  Figure~\ref{fig:minefield-semantics}.
\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
  e_f$ and $LF$, we can construct a liveness environment as follows:
  \begin{enumerate}
  \item If $e$ is an expression, then:  
  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
    \text{where }
  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
\item If $e$ is an application, then:  
  $$ \Lv = ref(e, \sigma, \Lfonly)$$
  \end{enumerate}
We can define in the same way the liveness environment for each of the
suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
environments that we shall denote $\mathsf{SL}$.
\item $GC$ models a liveness based garbage collection:
$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
  \in domain(\rho)$,  $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv'$.
\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \in \Lv$, .
  \end{enumerate}
\end{enumerate}

@@

\begin{proof}
For a program  executing according to {\em minefield  semantics}, if a
\bang\ has to  occur then some variable which is  dereferenced must be
set to  $\bot$, between its  creation and  use points.  This  can only
happen if  the liveness of  the variable being dereferenced  was null.
From  the lemma~\ref{lem:correctness}  it  is clear  that this  cannot
happen.
\end{proof}

@@

\balance
\bibliography{fun_hra}{}
\bibliographystyle{abbrv}

\appendix
\newcommand{\cred}[1]{{\color{red}{#1}}}
\section{Minefield semantics for the language}

>>>>

\begin{figure}[t]
\renewcommand{\arraystretch}{1}
\begin{eqnarray*}
   p \in \mathit{Prog} & \!\!\!::=\!\!\! & d_1 \ldots d_n \,\, e_\mainpgm
    \hspace{5em} \mbox{\em --- program}\\
    \mathit{df} \in Fdef & \!\!\!::=\!\!\! & (\DEFINE\,\, (f\,\, x_1 \,\, \ldots
\,\,x_n)\,\,
    e)
    \hspace{0.2em} \ \ \ \ \mbox{\em --- function def} \\
e \in \mathit{Expr} & \!\!\!::=\!\!\! &
\left\{\begin{array}{@{}ll@{\hspace{1em}}l}
       (\SIF\,\, x\,\, e_1\,\, e_2) && \!\!\!\mbox{\em --- conditional} \\
       (\LET\,\, x \leftarrow s\,\, \IN\,\, e) &&\!\!\! \mbox{\em --- let
binding} \\
       (\SRETURN\,\, x) && \!\!\!\mbox{\em --- return from function}
    \end{array}\right. \\
s \in \mathit{App} & \!\!\!::=\!\!\!  &
\left\{\begin{array}{@{}l@{\hspace{1.2em}}l}
       k & \mbox{\em --- constant (numeric or $\NIL$)}\\
       (\CONS\,\, x_1\,\, x_2) & \mbox{\em --- constructor} \\
       (\CAR\,\, x) & \\
       (\CDR\,\, x) & \mbox{\em --- selectors} \\
       (\NULLQ\,\, x) & \\
       (\PRIM\,\, x_1\,\, x_2) & \mbox{\em ---  tester/generic arithmetic} \\
%%       (\ID\,\, x) & \mbox{\em ---  identity function (for inlining)} \\
       (f\,\, x_1\,\,\ldots\,\, x_n) & \mbox{\em --- function application}
    \end{array}\right.
\end{eqnarray*}
  \caption{The syntax of our language}\label{fig:lang-syntax}
\figrule
\normalsize
\end{figure}

\begin{figure*}[t!]
\begin{center}
\renewcommand{\arraystretch}{1}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\
\hline
\hline
          & $\rho, (\rho', x, e)\!:\!S, H, \kappa
  \rightsquigarrow \rho', S, H[\rho'(x) := \kappa], e$    &  \sc{const}
\\
\hline
          & {$\rho, (\rho', z, e)\!:\!S, H, (\CONS~x~y)
\rightsquigarrow
$  $\rho', S, H[\rho'(z) := (\rho(x), \rho(y))], e$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & $\rho, (\rho', z, e)\!:\!S, H,
(\CAR~x)  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e$      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (\CAR~x)
\rightsquigarrow
\rho', (\rho, x, (\CAR~x))\!:\!S, H, s$      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & {$\rho, (\rho', z, e)\!:\!S, H, (+~x~y)  \rightsquigarrow$
$\rho', S, H[\rho'(z) := H(\rho'(x)) + H(\rho'(y))], e$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, x, (+~x~y))\!:\!S, H, s$      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & $\rho, S, H, (+~x~y)
\rightsquigarrow
\rho', (\rho, y, (+~x~y))\!:\!S, H, s$      &
\sc{prim-2-clo} \\
\hline
{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & $\rho, S, H,
(f~\myvec{x})  \rightsquigarrow
[\myvec{y} \mapsto \rho(\myvec{x})], S, H, e_{\mathit{f}}$      &
\sc{funcall} \\
\hline
$\ell$ is a new location& {$\rho, S, H, (\LET~x\leftarrow s~\IN~e)
  \rightsquigarrow$
$\rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s,
    \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_1$ & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & $\rho, S, H, (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho, S, H,  e_2$ & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H,
  (\SIF~x~e_1~e_2)   \rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2))\!:\!S, H,  s$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& $\rho, (\rho', z,
e)\!:\!S, H,
(\SRETURN~x)  \rightsquigarrow \rho', S, H[\rho'(z) := v], e$ &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & {$\rho, S, H, (\SRETURN~x)
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x))\!:\!S, H,  s$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{A small-step semantics for the language. \label{fig:lang-semantics}}
\end{center}
\end{figure*}

\subsection{Semantics}
We now give  a small-step semantics for our  language.
We first specify the domains used by the semantics:
\[
\renewcommand{\arraystretch}{1}
\begin{array}{@{}r@{\ }l@{\ \ }c@{\ \ }l@{\hspace{0.5em}}l}
\rho: & \mathit{Env} &=&\mathit{Var} \rightarrow \mathit{Loc} & \mbox{-- Environment} \\
v:   & \mathit{Val} &=& \mathbb{N} + \{\NIL\} + \mathit{Data \times Data}& \mbox{-- Values}\\
c:   & \mathit{Clo} &=& \mathit{(Exp \times Env)}& \mbox{-- Closures}\\
d: & \mathit{Data} &=&\mathit{Val} + \mathit{Clo} & \mbox{-- Values \& Closures} \\
\heap: & \mathit{Heap} & =&\mathit{Loc} \rightarrow Data & \mbox{-- Heap}
\end{array}
\]

@@

%==============================================================
\renewcommand{\pp}[2]{\ensuremath{#1\!\!:\!#2}} % prog point
%==============================================================
\begin{figure*}[t]
\renewcommand{\arraystretch}{1}
\begin{eqnarray*}
\mathit{ref\/}(\kappa,\sigma,\Lfonly)
          &=& \{\,\} \mbox{, for $\kappa$ a constant, including
$\NIL$}\\
\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
          &=& \{x.\alpha \mid \acar\alpha \in \sigma\} \cup \{y.\alpha
\mid \acdr\alpha \in \sigma\} \\
\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acar\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\CDR~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\} \cup \{x.\acdr\alpha \mid \alpha \in
\sigma\}, & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\PRIM~x~y),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon, y.\epsilon\},  & \mbox{if}~\sigma \ne
\emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((\NULLQ~x),\sigma,\Lfonly)
          &=&    \begin{array}{l l}
                    \{x.\epsilon\},  & \mbox{if}~\sigma \ne \emptyset\\
                    \emptyset  & \mbox{otherwise}
                 \end{array} \\
\mathit{ref\/}((f~\myvec{x}),\sigma,\Lfonly)
%          &=& \bigcup_{i=1}^n y_i.\Lf{f}{i}{\sigma}
          &=&  \begin{array}{@{}l}  % to discourage \displaystyle
               \bigcup_{i=1}^n x_i.\Lf{f}{i}{\sigma}
               \end{array}
%          &=& \bigcup \{y_i.\Lf{f}{i}{\sigma} \mid i=1,\ldots, n\}
\\[1ex]
\mathcal{L}((\SRETURN~\psi:x),\sigma,\Lfonly) &=& \lbrack \mathit{\epath}
\mapsto
x.\sigma\rbrack, \mbox{ where $\mathit{\epath}$ is a
  new e-path terminating with $\psi$}\\
\mathcal{L}((\SIF~\psi:x~e_1~e_2),\sigma,\Lfonly) &=&
\begin{array}{l l}
                    \mathcal{L}(e_1,\sigma,\Lfonly) \uplus
        \mathcal{L}(e_2,\sigma,\Lfonly) \uplus
        \lbrack\lbrace \mathit{\epath} \mapsto
\{x.\epsilon\}\rbrace\rbrack,  & \mbox{if}~\sigma \ne \emptyset\\
        \emptyset_{\cal{M}}  & \mbox{otherwise}
                 \end{array} \\
&&\mbox{where $\mathit{\epath}$ is a new e-path
  terminating with $\psi$}\\
\mathcal{L}(\LET~x \leftarrow~s~\IN~e),\sigma,\Lfonly) &=&
        \lambda\mathit{\epath}.\; \Lfonly_f(\Lv(x)) \cup \Lv~~
\mbox{where}~\mathcal{M} =
\mathcal{L}(e,\sigma,\Lfonly),~\Lv =
\mathcal{M}(\epath),~\mbox{and}~f~\mbox{is the new function}\\
&& \makebox[0mm]{\hspace*{7cm}
 $(\DEFINE~(f\,x_1\,x_2\, \ldots \, x_n)~(\SIF\,*~x~s[(f\,x_1\,
           x_2\, \ldots\, x_n)/x]))$,} \\
 \makebox[0mm]{\hspace*{8cm} where
     $x_1\, \ldots\, x_n$ are the variables in $s$}
\end{eqnarray*}
\begin{minipage}{0.85\textwidth}
\infrule[live-define]
        {\mathcal{L}(e_f,\sigma,\Lfonly) =
           \bigcup_{i=1}^n z_i.\Lf{f}{i}{\sigma}
              \mbox{ for each $f$ and $\sigma$}
        }
        { \mathit{df_1} \ldots \mathit{df_k} \len \Lfonly
\\ \makebox[0mm]{where
     $(\DEFINE\ (f\ z_1\ \ldots\ z_n)\ \ e_f)$ is a member of $\mathit{df_1}
\ldots \mathit{df_k}$}}
\end{minipage}
  \caption{Liveness equations and judgement rule}\label{fig:live-judge}
\end{figure*}

@@

\newcommand{\cred}[1]{{\color{red}{#1}}}
\section{Minefield semantics for the language}
\begin{figure*}[ht!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Premise & Transition & Rule name \\ 
\hline
\hline
          &\makecell{ $\rho, (\rho', x, e, \cred{\sigma', e_f'})\!:\!S,
  H, \kappa, \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(x) :=
    \kappa], e, \cred{\sigma', e_f'}$ }   &  \sc{const}
\\
\hline
          & \makecell{$\rho, (\rho', z, e, \cred{\sigma',
    e_f'})\!:\!S, H, (\CONS~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
  \rho', S, H[\rho'(z) := (\rho(x),\rho(y))], e, \cred{\sigma', e_f'}$}     &  \sc{cons} \\
\hline
$H(\rho(x)) \mbox{ is } (l_1, l_2)$ & \makecell{$\rho, (\rho', z, e,
  \cred{\sigma', e_f'} )\!:\!S, H, (\CAR~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow \rho', S, H[\rho'(z) := H(l_1)], e, \cred{\sigma', e_f'}$}      &
\sc{car-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{ $\rho, S,
  H, (\CAR~x), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (\CAR~x), \cred{\sigma, e_f})\!:\!S, H, s, \cred{(\clazy \cup \acar)\sigma , \_ }$}      &
\sc{car-clo}
\\
\hline
$H(\rho(x)), H(\rho(y)) \in \mathbb{N}$
 & \makecell{$\rho, (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H,
  (+~x~y), \cred{\sigma, e_f}$ \\  $\rightsquigarrow \rho', S, H[\rho'(z)
    := H(\rho'(x)) + H(\rho'(y))], e, \cred{\sigma', e_f'}$}      &
\sc{prim-whnf} \\
\hline
$H(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$ &\makecell{$\rho, S,
  H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, x,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_}$}      &
\sc{prim-1-clo} \\
\hline
$H(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $ & \makecell{$\rho,
  S, H, (+~x~y), \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho', (\rho, y,
  (+~x~y), \cred{\sigma, e_f})\!:\!S, H, s, \cred{\clazy\sigma, \_ }$}      &
\sc{prim-2-clo} \\
\hline
{$\mathit{g}~\mbox{defined as}$
$~(\DEFINE~(g~\myvec{y})~e_{\mathit{g}})$}  & \makecell{$\rho, S, H,
  (g~\myvec{x}), \cred{\sigma, e_f}$ \\ $\rightsquigarrow [\myvec{y} \mapsto
    \rho(\myvec{x})], S, H, e_{\mathit{g}}, \cred{\sigma, e_g}$}      &
\sc{funcall} \\
\hline
$\ell$ is a new location& \makecell{$\rho, S, H, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma, e_f}$ \\ $ \rightsquigarrow \rho\oplus[x \mapsto \ell], S, H[\ell \mapsto \langle s, \lfloor\rho\rfloor_{FV(s)}  \oplus [x \mapsto
  \ell]\rangle], e, \cred{\sigma, e_f}$} &
\sc{let} \\
\hline
$H(\rho(x)) \ne 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\ $\rightsquigarrow \rho, S, H,  e_1, \cred{\sigma, e_f}$} & \sc{if-true} \\
\hline
$H(\rho(x)) = 0$ & \makecell{$\rho, S, H, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma, e_f}$ \\   $\rightsquigarrow
\rho, S, H,  e_2, \cred{\sigma, e_f}$} & \sc{if-false} \\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma, e_f}$ \\ $\rightsquigarrow
\rho', (\rho, x, (\SIF~x~e_1~e_2),  \cred{\sigma, e_f})\!:\!S, H, s,
\cred{\clazy\sigma, e_f}$}
&
\sc{if-clo} \\
\hline
{$H(\rho(x))~\mbox{is}$ $\mbox{whnf with value}~v$}& \makecell{$\rho,
  (\rho', z, e, \cred{\sigma', e_f'})\!:\!S, H, (\SRETURN~x), \cred{\sigma,
  e_f}$ \\ $\rightsquigarrow \rho', S, H[\rho'(z) := v], e,
  \cred{\sigma', e_f'}$} &
\sc{return-whnf}\\
\hline
$H(\rho(x)) = \langle s, \rho' \rangle $ & \makecell{$\rho, S, H,
  (\psi:\SRETURN~x), \cred{\sigma, e_f}$ \\ $
  \rightsquigarrow$
$\rho',~ (\rho, x, (\SRETURN~x), \cred{\sigma, e_f})\!:\!S, H,  s,
  \cred{\sigma, e_f}$} &
\sc{return-clo} \\
\hline
\end{tabular}
\caption{Minefield semantics.\label{fig:minefield-semantics}}
\end{center}
\end{figure*}

To set up the minefield semantics, we follow these steps:
\begin{enumerate}
\item We enrich  the abstract machine state $\rho,$ $S,$  $ H,$ $e$ to
  $\rho,$ $S,$ $H,$ $e,$ $\sigma,$ $e_f$, where $\sigma$ is the demand
  on the expression $e$, and $e_f$ is the body of the function. As the
  section on liveness  analysis shows, the entire  function body $e_f$
  and $\sigma$  are needed  to calculate  the liveness  environment at
  different program  points. The  demand sigma  is ``dynamic''  in the
  sense  that it  arises from  the actual  function calls  made during
  execution.   The  0-CFA  demand  is an  over-approximation  of  this
  demand.  The information  in a suspended evaluation  context is also
  similarly augmented with  the demand on it and the  function body of
  which the suspended evaluation is a  part.  Thus a stacked entry now
  takes the  form $(\rho, x, e,  \sigma, e_f)$. A modification  of the
  small step semantics which carries the extra information is shown in
  Figure~\ref{fig:minefield-semantics}.
\item Given  the current  evaluation context $\rho,  S, H,  e, \sigma,
  e_f$ and $LF$, we can construct a liveness environment as follows:
  \begin{enumerate}
  \item If $e$ is an expression, then:  
  $$ \Lv = \bigcup_{\psi \in \mathcal{P}(e)} \mathcal{M}(\psi),
    \text{where }
  \mathcal{M} = \mathcal{L}(e_f, \sigma, \Lfonly)$$
\item If $e$ is an application, then:  
  $$ \Lv = ref(e, \sigma, \Lfonly)$$
  \end{enumerate}
We can define in the same way the liveness environment for each of the
suspended  evaluation contexts in  $S$  giving  a  stack of  liveness
environments that we shall denote $\mathsf{SL}$.
\item $GC$ models a liveness based garbage collection:
$GC(\rho,$ $ S,$ $ H,$ $ L,$ $ SL) = (\rho', S', H')$ 
  \begin{enumerate}
  \item For each $x \in domain(\rho)$,  $\rho'(x)=\bot$ iff $x.\epsilon
    \notin \Lv$.
\item For each stack entry $(\rho,\_,\_,\_,\_)$ in $S$ with $\Lv'$ as
  the corresponding liveness environment in $\mathsf{SL}$,  and for each $x
  \in domain(\rho)$,  $\rho'(x)=\bot$
  iff $x.\epsilon \notin \Lv'$.
\item For each location $\ell$, $H'(\ell) = \bot$, iff  there is no
  $x$ in either the current environment or one of the stacked
  environments $\rho$ such that for some $\alpha$, $H[x.\alpha] =
  \ell$  and $x.\alpha \in \Lv$, .
  \end{enumerate}
\end{enumerate}

@@

\begin{proof}
For a program  executing according to {\em minefield  semantics}, if a
\bang\ has to  occur then some variable which is  dereferenced must be
set to  $\bot$, between its  creation and  use points.  This  can only
happen if  the liveness of  the variable being dereferenced  was null.
From  the lemma~\ref{lemna:no-calls-correctness}  it  is clear  that this  cannot
happen.
\end{proof}

@@

\balance
\bibliography{fun_hra}{}
\bibliographystyle{abbrv}
\end{document}
